\chapter{To-Do List}\label{ch:todo}

\textbf{Potential committee members:}
\begin{itemize}
    \item \textbf{Anton Kaul}
    \item \textbf{Patrick Orson}
    \item \textbf{Eric Brussel}
    \item \textit{Rob Easton}
    % \item Tony Mendes (not likely)
    % \item Emily Hamilton
    % \item Jeffrey Liese
    % \item Ben Richert
    % \item Dana Paquin (not here?)
    
    % \textit{Physics Faculty:}
    % \item Thomas Gutierrez (Quantum info sci.)
    % \item Hilary Jacks (Two-level memory systems)
    % \item Lei Lu (Non-commutative (quantum?) field theory)
    % \item Matt Mewes (Theoretical particle physics)
    % \item Ian Powell (CMT)
    % \item Karl Saunders (Soft CMT)
    % \item Ben Shlaer (Th. particle, string, cosmo.)
\end{itemize}

\begin{center}\rule{.85\textwidth}{0.65pt}\end{center}

\begin{itemize}

    \item \textbf{First:} Do Lorentz group example, finish off content that you want in there. Then go back and fix notation once you know that you won't need it for Lorentz group. Alternatively, you can just invent the notation solely for that example!
    \item Redo the \cref{ch:rep_background} with nicer notation and stray away from Tung's notation when possible.
    \item More straightforward examples of representations?
    \item At least briefly discuss $U(n)$ either here or in braid rep chapter.
    \item Finish/modify irreducible rep. example in \cref{ch:rep_background}.
    \item Do more for $\textrm{SO}(2)$ example?
    \item Fix out equation numbers
    \item \textcolor{Green}{\textbf{Examples to add:}} 1D conservation of momentum from $T$, ang-momentum connection to SO(2)?, \sout{Lorentz group}
    \item Do a little more context on the physics. Maybe put in appendix, but describe what the heck a quantum Hilbert space is and braket notation.
    \item Also note why we care about unitary matrices, probs in appendix as well.
    
    \begin{center}\rule{.85\textwidth}{0.65pt}\end{center}
    
    \item Show $\psi_n(\sigma_i)$ invertible? Yes, eventually
    \item derive $\psi_n^\textbf{r}(\sigma_i)$ matrices or state?
    \item Show $\psi_n^\textbf{r}(\sigma_i)$ invertible? Yes, eventually
    \item \sout{Explicitly show why Burau isn't able to be made unitary?~\cite{Delaney2016}}
    \item Separate chapters into braid group and braid group reps.?
    
    \begin{center}\rule{.85\textwidth}{0.65pt}\end{center}
    
    \item Concluding paragraph on first section to lead into the more physics-y stuff.
    \item \sout{Show the additional cross terms from $N=2$ to $N=3$ and beyond.}
    \item Add paragraph on gauge theory/motivation.
    \item Anyon fusion rules
    \item $\tau$ anyon/Fibonacci anyon example. Relate to singlet/triplet states in spin-1/2 system.
    \item \sout{Move anyon calculations to appendix?}
    \item Spend some time on \texttt{MATLAB} thing

    \begin{center}\rule{.85\textwidth}{0.65pt}\end{center}
    
    \item Conclusion/future of anyons/braid group in physics.
    \item Abstract
    \item Title
    \item Acknowledgements
    
    \begin{center}\rule{.85\textwidth}{0.65pt}\end{center}
    
    \textbf{Format!!}
\end{itemize}
